%
% gameplay.tex
%
% Crypt: Gameplay Mechanics Specification
% Defines all game rules as implemented in internal/engine/.
%
% Compile: pdflatex gameplay.tex
%

\documentclass[a4paper,10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{longtable}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  framerule=0.3pt,
  rulecolor=\color{gray!50},
  backgroundcolor=\color{gray!5},
  columns=fullflexible,
}

\newcommand{\dice}[1]{\texttt{#1}}
\newcommand{\stat}[1]{\textsc{#1}}
\newcommand{\cmd}[1]{\texttt{#1}}
\newcommand{\field}[1]{\texttt{#1}}

\begin{document}

\title{Crypt: Gameplay Mechanics Specification}
\author{Punt Labs}
\date{March 2026 --- v1.0}
\hypersetup{
  pdftitle={Crypt: Gameplay Mechanics Specification},
  pdfauthor={Punt Labs},
  pdfsubject={Game Design},
  pdfcreator={pdflatex}
}
\maketitle
\tableofcontents
\newpage

% ===================================================================
\section{Design Principles}
% ===================================================================

All game rules are implemented deterministically in \texttt{internal/engine/}.
The engine is the single source of truth for state transitions. No game rule
or state mutation is left to probabilistic LLM output (DES-009).

\begin{itemize}[nosep]
  \item All randomness uses dice notation: \dice{NdS} (roll $N$ dice with
        $S$ sides) with optional modifiers \dice{NdS+M} or \dice{NdS-M}.
  \item All damage floors at 1 --- no attack can deal 0 damage.
  \item All HP caps at \field{MaxHP} --- healing cannot exceed the maximum.
  \item The engine uses \texttt{math/rand} for all dice rolls. Seeding the
        global RNG before a session produces a fully deterministic playthrough.
\end{itemize}

% ===================================================================
\section{Character Model}
% ===================================================================

% -------------------------------------------------------------------
\subsection{Attributes}
% -------------------------------------------------------------------

Every character has six attributes, inspired by Wizardry~I (DES-022).

\begin{table}[h]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Attribute} & \textbf{Abbreviation} & \textbf{Default} & \textbf{Mechanical Effect} \\
\midrule
Strength     & \stat{STR} & 14 & None (reserved for damage bonus) \\
Dexterity    & \stat{DEX} & 12 & Initiative roll, flee success rate \\
Constitution & \stat{CON} & 12 & HP bonus per level: $(\stat{CON} - 10) / 2$ \\
Intelligence & \stat{INT} & 10 & None (reserved for spell power) \\
Wisdom       & \stat{WIS} & 10 & None (reserved for MP regen) \\
Charisma     & \stat{CHA} & 10 & None (reserved for shop prices) \\
\bottomrule
\end{tabular}
\caption{Character attributes. All characters start with identical values.
         Only \stat{DEX} has a mechanical effect in the current implementation.}
\label{tab:attributes}
\end{table}

% -------------------------------------------------------------------
\subsection{Classes}
% -------------------------------------------------------------------

Four classes are available. All share the same starting attributes and HP
(20). Classes differ in MP access, XP progression rate, and per-level gains.

\begin{table}[h]
\centering
\begin{tabular}{@{}lccccl@{}}
\toprule
\textbf{Class} & \textbf{Starting MP} & \textbf{HP/Level} & \textbf{MP/Level} & \textbf{Can Cast} & \textbf{Primary Stats} \\
\midrule
Fighter & 0  & +8 & +0 & No  & \stat{STR}, \stat{CON} \\
Mage    & 10 & +4 & +4 & Yes & \stat{INT}, \stat{WIS} \\
Priest  & 10 & +6 & +3 & Yes & \stat{WIS}, \stat{CHA} \\
Thief   & 0  & +6 & +0 & No  & \stat{DEX}, \stat{CHA} \\
\bottomrule
\end{tabular}
\caption{Class progression. Primary stats gain +1 per level-up.}
\label{tab:classes}
\end{table}

% -------------------------------------------------------------------
\subsection{Starting State}
% -------------------------------------------------------------------

All new characters begin with:

\begin{table}[h]
\centering
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Field} & \textbf{Value} \\
\midrule
Level & 1 \\
HP / MaxHP & 20 / 20 \\
MP / MaxMP & 10 / 10 (mage, priest) or 0 / 0 (fighter, thief) \\
Gold & 0 \\
Inventory & Empty \\
Equipment & All slots empty (weapon, armor, ring, amulet) \\
Stats & \stat{STR}~14, \stat{DEX}~12, \stat{CON}~12, \stat{INT}~10, \stat{WIS}~10, \stat{CHA}~10 \\
\bottomrule
\end{tabular}
\caption{Default character creation values. No point-buy or random rolling.}
\end{table}

% ===================================================================
\section{Experience and Leveling}
% ===================================================================

% -------------------------------------------------------------------
\subsection{XP Awards}
% -------------------------------------------------------------------

XP is awarded when an enemy is killed. The award equals the enemy's
\field{MaxHP}, not its current HP at death.

\[
  \text{XP}_\text{award} = \text{enemy.MaxHP}
\]

XP is cumulative and never lost. Level-up checks occur after each combat
victory (all enemies defeated). Multiple level-ups can occur in a single
check if accumulated XP is sufficient.

% -------------------------------------------------------------------
\subsection{XP Thresholds}
% -------------------------------------------------------------------

Each class has a distinct XP table. Fighters level cheaply; mages require
the most XP. Inspired by Wizardry~I's class-based progression rates.

\begin{table}[h]
\centering
\begin{tabular}{@{}rrrrr@{}}
\toprule
\textbf{Level} & \textbf{Fighter} & \textbf{Thief} & \textbf{Priest} & \textbf{Mage} \\
\midrule
2  & 20    & 22    & 25    & 30    \\
3  & 50    & 55    & 60    & 75    \\
4  & 100   & 110   & 120   & 150   \\
5  & 200   & 220   & 250   & 300   \\
6  & 400   & 440   & 500   & 600   \\
7  & 800   & 880   & 1000  & 1200  \\
8  & 1600  & 1760  & 2000  & 2400  \\
9  & 3200  & 3520  & 4000  & 4800  \\
10 & 6400  & 7040  & 8000  & 9600  \\
\bottomrule
\end{tabular}
\caption{Cumulative XP required to reach each level. Max level is 10.}
\label{tab:xp}
\end{table}

% -------------------------------------------------------------------
\subsection{Level-Up Effects}
% -------------------------------------------------------------------

On level-up, effects are applied in this order:

\begin{enumerate}[nosep]
  \item \textbf{Stats:} +1 to each of the class's two primary stats
        (see Table~\ref{tab:classes}). Applied first so the new \stat{CON}
        value is used for the HP calculation.
  \item \textbf{HP:} class base amount + \stat{CON} modifier, minimum 1:
        \[
          \text{HP gain} = \max\!\big(1,\; \text{hpPerLevel}[\text{class}] + \lfloor(\stat{CON} - 10) / 2\rfloor\big)
        \]
        Fighters (base 8, CON growth) gain 9--10+ HP at higher levels.
        Mages (base 4, no CON growth) gain 4 HP throughout.
  \item \textbf{MP:} class-specific amount added to both \field{MP} and
        \field{MaxMP} (fighters and thieves gain 0).
\end{enumerate}

% ===================================================================
\section{Combat}
% ===================================================================

Combat is turn-based and auto-starts when the hero enters a room containing
enemies (unless the room has been cleared by a previous victory).

% -------------------------------------------------------------------
\subsection{Initiative}
% -------------------------------------------------------------------

At combat start, initiative determines turn order:

\begin{align}
  \text{Hero initiative}  &= \stat{DEX} + \dice{1d20} \\
  \text{Enemy initiative} &= 10 + \dice{1d20} \quad \text{(per enemy)}
\end{align}

Turn order is sorted descending by initiative roll. Ties are broken in
favor of the hero. Once set, turn order is fixed for the duration of
combat. Dead enemies are skipped during turn processing.

% -------------------------------------------------------------------
\subsection{Hero Actions}
% -------------------------------------------------------------------

On the hero's turn, one of four actions may be taken:

\subsubsection{Attack}

The hero attacks a target enemy (defaults to the first alive enemy if no
target is specified).

\[
  \text{damage} = \begin{cases}
    \text{Roll}(\text{weapon.Damage}) & \text{if weapon equipped} \\
    \dice{1d2} & \text{unarmed}
  \end{cases}
\]

Damage is applied directly to the target's HP. Minimum damage is 1.
If the target's HP drops to 0 or below, the target is killed and the hero
is awarded XP equal to the target's \field{MaxHP}. If all enemies are dead,
combat ends and the room is marked \field{Cleared}.

\subsubsection{Defend}

The hero assumes a defensive stance. Incoming damage from all enemies is
halved (integer division, floor 1) until the hero's next turn. The
\field{HeroDefending} flag is cleared when the hero's turn comes around
again.

\subsubsection{Flee}

The hero attempts to escape combat.

\[
  \text{Roll } \dice{1d20}: \quad \text{success if roll} \leq \stat{DEX}
\]

With the default \stat{DEX} of 12, the base flee rate is $12/20 = 60\%$.
Thieves gain +1 \stat{DEX} per level, improving their flee rate.

\begin{itemize}[nosep]
  \item \textbf{Success:} Combat ends. The room is \emph{not} marked cleared
        --- enemies will attack again if the hero re-enters.
  \item \textbf{Failure:} The hero loses their turn. Enemies still act.
\end{itemize}

\subsubsection{Cast Spell}

See Section~\ref{sec:spells}.

% -------------------------------------------------------------------
\subsection{Enemy Actions}
% -------------------------------------------------------------------

After the hero acts, all enemies take their turns in initiative order
before the hero acts again. Each enemy follows its AI pattern:

\begin{table}[h]
\centering
\begin{tabular}{@{}lp{9cm}@{}}
\toprule
\textbf{AI} & \textbf{Behavior} \\
\midrule
\texttt{aggressive} & Always attacks. \\
\texttt{cautious}   & Attacks normally. Flees (removed from combat) when
                       $\text{HP} \leq 30\%$ of \field{MaxHP}. Flee awards
                       no XP. \\
\texttt{scripted}    & Alias for aggressive. \\
\bottomrule
\end{tabular}
\caption{Enemy AI patterns.}
\end{table}

Enemy attack damage:

\[
  \text{raw damage} = \text{Roll}(\text{enemy.Attack})
\]

Damage is then reduced by defenses (see Section~\ref{sec:armor}).

% -------------------------------------------------------------------
\subsection{Damage Reduction}
\label{sec:armor}
% -------------------------------------------------------------------

When an enemy attacks the hero, raw damage is reduced by two mechanisms
applied in order:

\begin{enumerate}[nosep]
  \item \textbf{Defend stance:} If the hero is defending,
        $\text{damage} = \lfloor \text{damage} / 2 \rfloor$.
  \item \textbf{Armor defense:} If the hero has armor equipped with a
        \field{defense} value $d > 0$, $\text{damage} = \text{damage} - d$.
\end{enumerate}

After both reductions, damage is floored at 1:

\[
  \text{final damage} = \max\!\big(1,\; \lfloor \text{raw} / k \rfloor - d\big)
  \quad \text{where } k = \begin{cases} 2 & \text{defending} \\ 1 & \text{otherwise} \end{cases}
\]

% -------------------------------------------------------------------
\subsection{Death}
% -------------------------------------------------------------------

When the hero's HP drops to 0 or below during enemy turns, the hero dies.
The game loop terminates immediately after narrating the death event.
The scenario's \field{death} field (\texttt{respawn} or \texttt{permadeath})
determines whether the game can be reloaded, but the engine does not
currently implement respawn mechanics.

% -------------------------------------------------------------------
\subsection{Room Clearing}
% -------------------------------------------------------------------

When all enemies in a room are killed (HP~$\leq 0$), the room's
\field{Cleared} flag is set to \texttt{true}. A cleared room will not
trigger combat on re-entry. Enemies that flee (cautious AI) are removed
from combat; if all remaining enemies are dead or fled, the room is cleared
and the hero is awarded XP for kills (not for fled enemies).

% ===================================================================
\section{Spells}
\label{sec:spells}
% ===================================================================

Spells are defined in scenario YAML with five fields: \field{name},
\field{mp} (cost), \field{effect} (\texttt{damage} or \texttt{heal}),
\field{power} (dice notation), and \field{classes} (list of allowed classes).

% -------------------------------------------------------------------
\subsection{Casting Prerequisites}
% -------------------------------------------------------------------

A spell cast fails (and costs no MP) if any of the following are true:

\begin{enumerate}[nosep]
  \item The hero is dead (HP~$\leq 0$).
  \item The spell ID does not exist in the scenario.
  \item The hero's class is not in the spell's \field{classes} list.
  \item The hero's MP is less than the spell's \field{mp} cost.
  \item (Damage spells only) Combat is not active.
  \item (Damage spells only) It is not the hero's turn.
  \item (Damage spells only) The target is invalid (dead or nonexistent).
\end{enumerate}

MP is deducted \emph{after} the spell effect resolves, never on failure.

% -------------------------------------------------------------------
\subsection{Damage Spells}
% -------------------------------------------------------------------

Damage spells require active combat and a valid target. The spell's
\field{power} is rolled as dice notation.

\[
  \text{spell damage} = \max(1,\; \text{Roll}(\text{spell.Power}))
\]

The damage is applied directly to the target's HP. If the target dies,
XP is awarded (equal to target's \field{MaxHP}) and combat may end.
Casting a damage spell consumes the hero's turn.

% -------------------------------------------------------------------
\subsection{Heal Spells}
% -------------------------------------------------------------------

Heal spells work both in and out of combat. When cast in combat, they
consume the hero's turn. The spell's \field{power} is rolled as dice
notation and added to the hero's HP, capped at \field{MaxHP}.

\[
  \text{hero.HP} = \min(\text{MaxHP},\; \text{hero.HP} + \max(1,\; \text{Roll}(\text{spell.Power})))
\]

% ===================================================================
\section{Inventory and Equipment}
% ===================================================================

% -------------------------------------------------------------------
\subsection{Item Types}
% -------------------------------------------------------------------

Items have a \field{type} field that determines how they can be used:

\begin{table}[h]
\centering
\begin{tabular}{@{}llp{7cm}@{}}
\toprule
\textbf{Type} & \textbf{Slot} & \textbf{Behavior} \\
\midrule
\texttt{weapon}     & Weapon & Equipped to weapon slot. \field{damage} field used for attack rolls. \\
\texttt{armor}      & Armor  & Equipped to armor slot. \field{defense} field reduces incoming damage. \\
\texttt{ring}       & Ring   & Equipped to ring slot. No mechanical effect (reserved). \\
\texttt{amulet}     & Amulet & Equipped to amulet slot. No mechanical effect (reserved). \\
\texttt{consumable} & ---    & Used via \cmd{use} command. Consumed on use. \field{effect} and \field{power} determine the effect. \\
\texttt{key}        & ---    & Carried in inventory. No use action (reserved for locked doors). \\
\texttt{misc}       & ---    & Carried in inventory. No mechanical effect. \\
\bottomrule
\end{tabular}
\caption{Item types and their equipment behavior.}
\end{table}

% -------------------------------------------------------------------
\subsection{Weight Limit}
% -------------------------------------------------------------------

Every character has a carry weight limit of \textbf{50.0} units. Total
weight includes both inventory items and equipped items. Picking up an
item that would exceed the limit returns a \texttt{TooHeavyError}.

% -------------------------------------------------------------------
\subsection{Equipment Slots}
% -------------------------------------------------------------------

Four equipment slots exist: weapon, armor, ring, amulet. Each slot holds
at most one item. Equipping an item removes it from inventory and places
it in the slot. Unequipping moves it back to inventory. Attempting to equip
into an occupied slot returns a \texttt{SlotOccupiedError}.

% -------------------------------------------------------------------
\subsection{Consumable Items}
% -------------------------------------------------------------------

Consumable items are used via the \cmd{use} command. On use, the item is
removed from inventory and its effect is applied:

\begin{itemize}[nosep]
  \item \textbf{Heal:} Roll \field{power} dice, add to hero HP (capped
        at \field{MaxHP}). Can be used in or out of combat.
\end{itemize}

Using a non-consumable item returns a \texttt{NotConsumableError}.
Using consumables does not consume a combat turn (it is processed outside
the combat turn system).

% ===================================================================
\section{Movement and Exploration}
% ===================================================================

% -------------------------------------------------------------------
\subsection{Connections}
% -------------------------------------------------------------------

Rooms are connected by directional exits. Six directions are supported:
north, south, east, west, up, down. Each connection has a type:

\begin{table}[h]
\centering
\begin{tabular}{@{}lp{8cm}@{}}
\toprule
\textbf{Type} & \textbf{Behavior} \\
\midrule
\texttt{open}     & Freely traversable. \\
\texttt{locked}   & Returns \texttt{LockedError}. Cannot be traversed (key mechanics not yet implemented). \\
\texttt{hidden}   & Invisible to the player. Not listed in exits. Returns \texttt{NoExitError} if attempted. \\
\texttt{stairway} & Freely traversable (functionally identical to \texttt{open}). \\
\bottomrule
\end{tabular}
\caption{Connection types.}
\end{table}

% -------------------------------------------------------------------
\subsection{Room State}
% -------------------------------------------------------------------

Each room has mutable state tracking:

\begin{itemize}[nosep]
  \item \field{Cleared}: whether all enemies have been defeated.
        A cleared room does not trigger combat on re-entry.
  \item \field{Items}: list of item IDs currently in the room. Items can
        be picked up (removed from room, added to inventory) or dropped
        (added to room from inventory). Room items are seeded from the
        scenario on \texttt{NewGame}.
\end{itemize}

% -------------------------------------------------------------------
\subsection{Fog of War}
% -------------------------------------------------------------------

The \field{VisitedRooms} list tracks which rooms the hero has entered.
Rooms not in this list are unvisited. The list grows monotonically ---
rooms are never removed.

% -------------------------------------------------------------------
\subsection{Combat Auto-Start}
% -------------------------------------------------------------------

When the hero moves into a room that has enemies and is not cleared,
\texttt{StartCombat} is called automatically after the move resolves.
The hero sees the room description, then immediately enters combat.
If enemies win initiative, they act before the hero.

% ===================================================================
\section{Commands}
% ===================================================================

The \texttt{RulesInterpreter} maps text input to engine actions. All
commands are case-insensitive. Articles (\textit{the}, \textit{a},
\textit{an}) are stripped. Multi-word targets are joined with underscores.
Typos within edit distance~1 of known verbs are auto-corrected.

\begin{longtable}{@{}llp{6cm}@{}}
\toprule
\textbf{Input} & \textbf{Aliases} & \textbf{Action} \\
\midrule
\endfirsthead
\toprule
\textbf{Input} & \textbf{Aliases} & \textbf{Action} \\
\midrule
\endhead
\cmd{north}, \cmd{south}, etc. & \cmd{n}, \cmd{s}, \cmd{e}, \cmd{w}, \cmd{u}, \cmd{d} & Move in direction \\
\cmd{go <dir>}         & ---                        & Move in direction \\
\cmd{ascend}           & ---                        & Move up \\
\cmd{descend}          & ---                        & Move down \\
\cmd{look}             & \cmd{l}                    & Describe current room \\
\cmd{examine <item>}   & \cmd{x}                    & Inspect an item \\
\cmd{look at <item>}   & ---                        & Alias for examine \\
\cmd{take <item>}      & \cmd{get}, \cmd{grab}      & Pick up from room \\
\cmd{pick up <item>}   & ---                        & Pick up from room \\
\cmd{drop <item>}      & ---                        & Drop to room \\
\cmd{equip <item>}     & \cmd{wear}, \cmd{wield}    & Equip from inventory \\
\cmd{unequip <slot>}   & \cmd{remove}               & Unequip to inventory \\
\cmd{use <item>}       & \cmd{drink}, \cmd{eat}     & Use consumable item \\
\cmd{attack [target]}  & \cmd{a}, \cmd{hit}, \cmd{strike}, \cmd{kill} & Attack enemy (combat only) \\
\cmd{defend}           & \cmd{block}, \cmd{guard}   & Defensive stance (combat only) \\
\cmd{flee}             & \cmd{run}, \cmd{escape}    & Attempt to flee (combat only) \\
\cmd{cast <spell>}     & ---                        & Cast a spell \\
\cmd{cast <spell> at <target>} & ---                & Cast damage spell at target \\
\cmd{inventory}        & \cmd{i}                    & List inventory and equipment \\
\cmd{save [slot]}      & ---                        & Save game (default: quicksave) \\
\cmd{load [slot]}      & ---                        & Load game (default: quicksave) \\
\cmd{help}             & \cmd{?}                    & List available commands \\
\cmd{quit}             & \cmd{exit}, \cmd{q}        & End the game \\
\bottomrule
\end{longtable}

Actions blocked during combat: \cmd{move}, \cmd{take}, \cmd{drop},
\cmd{equip}, \cmd{unequip}, \cmd{examine}. Attempting them returns an
\texttt{in\_combat} event.

% ===================================================================
\section{Save and Load}
% ===================================================================

The game supports named save slots (default: \texttt{quicksave}). The
full \texttt{GameState} is serialized to JSON with a \field{schema\_version}
field. Unknown fields in save files are silently ignored for forward
compatibility. Loading replaces the entire in-memory game state.

Save files are stored in \texttt{.dungeon/saves/<slot>.json}.

% ===================================================================
\section{Scenario Format}
% ===================================================================

Scenarios are defined in YAML. Two formats are supported:

\begin{enumerate}[nosep]
  \item \textbf{Single-file:} one \texttt{.yaml} file with all rooms,
        items, enemies, and spells.
  \item \textbf{Directory:} a directory containing \texttt{scenario.yaml}
        (manifest with title, starting room, catalogs, and region list)
        plus \texttt{regions/*.yaml} sub-files. Room IDs are globally
        unique across regions.
\end{enumerate}

Scenario resolution (\texttt{scenariodir.Load}): tries directory format
first (\texttt{id/scenario.yaml}), falls back to single-file
(\texttt{id.yaml}).

% -------------------------------------------------------------------
\subsection{Scenario Fields}
% -------------------------------------------------------------------

\begin{table}[h]
\centering
\begin{tabular}{@{}llp{7cm}@{}}
\toprule
\textbf{Field} & \textbf{Type} & \textbf{Description} \\
\midrule
\field{title}         & string & Display name \\
\field{starting\_room} & string & Room ID where the adventure begins \\
\field{death}         & string & \texttt{respawn} or \texttt{permadeath} \\
\field{rooms}         & map    & Room definitions keyed by ID \\
\field{items}         & map    & Item templates keyed by ID \\
\field{enemies}       & map    & Enemy templates keyed by ID \\
\field{spells}        & map    & Spell templates keyed by ID \\
\bottomrule
\end{tabular}
\caption{Top-level scenario fields.}
\end{table}

% ===================================================================
\section{Balance Tuning}
% ===================================================================

The \texttt{eval-balance} tool runs Monte Carlo simulations using a
monkey-testing harness (\texttt{internal/monkeytest}). The monkey player
makes weighted-random decisions based on game state:

\begin{table}[h]
\centering
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{State} & \textbf{Attack} & \textbf{Defend} & \textbf{Flee} & \textbf{Heal} \\
\midrule
Combat, HP $> 30\%$  & 70\% & 15\% & 10\% & 5\%\textsuperscript{*} \\
Combat, HP $\leq 30\%$ & 20\% & 20\% & 40\% & 20\%\textsuperscript{*} \\
\bottomrule
\end{tabular}
\caption{Combat action weights. \textsuperscript{*}Heal only available to
         casters with sufficient MP; otherwise weight redistributed.}
\end{table}

Out of combat, the monkey auto-equips weapons and armor, uses health
potions at $\leq 50\%$ HP, picks up items (60\%), moves randomly (30\%),
or looks around (10\%).

% -------------------------------------------------------------------
\subsection{Key Metrics}
% -------------------------------------------------------------------

Per-session metrics collected:

\begin{multicols}{2}
\begin{itemize}[nosep]
  \item Total moves
  \item Rooms visited (unique)
  \item Enemies killed
  \item XP gained
  \item Final level
  \item Levels gained
  \item Items picked up / equipped
  \item Combat rounds fought
  \item Damage dealt / taken
  \item Survival (alive at end)
  \item Flee attempts / successes
  \item Spells cast / heals cast
  \item Potions used
\end{itemize}
\end{multicols}

Aggregate report includes: survival rate, mean/median/p95 for all metrics,
level-up rate, per-class breakdowns. Output is JSON to stdout.

% -------------------------------------------------------------------
\subsection{Design Targets}
% -------------------------------------------------------------------

Target ranges for a well-balanced scenario:

\begin{table}[h]
\centering
\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Metric} & \textbf{Target Range} \\
\midrule
Survival rate (all classes)          & 50--75\% \\
Class survival spread (max -- min)   & $< 20$ percentage points \\
Level-up rate                        & $> 50\%$ \\
Mean enemies killed                  & $\geq 3$ \\
Flee success rate                    & 55--70\% (consistent with \stat{DEX} 12) \\
\bottomrule
\end{tabular}
\caption{Balance targets for monkey testing. A scenario that falls outside
         these ranges needs tuning.}
\end{table}

% ===================================================================
\section{Notation Reference}
% ===================================================================

\begin{table}[h]
\centering
\begin{tabular}{@{}lp{8cm}@{}}
\toprule
\textbf{Notation} & \textbf{Meaning} \\
\midrule
\dice{NdS}   & Roll $N$ dice, each with $S$ sides. Sum the results. \\
\dice{NdS+M} & Roll $N$d$S$ and add modifier $M$. \\
\dice{NdS-M} & Roll $N$d$S$ and subtract modifier $M$. \\
\dice{1d20}  & Standard 20-sided die roll (range 1--20). \\
\dice{1d2}   & Unarmed damage (range 1--2). \\
\dice{1d6}   & Standard weapon damage (short sword). \\
\dice{2d6}   & Spell damage (fireball). \\
\dice{1d6+2} & Spell heal (heal spell). \\
\bottomrule
\end{tabular}
\caption{Dice notation used throughout the engine.}
\end{table}

% ===================================================================
\section{Unimplemented Mechanics}
\label{sec:unimplemented}
% ===================================================================

The following mechanics are defined in the data model but have no
engine implementation:

\begin{enumerate}[nosep]
  \item \stat{STR} modifier to melee damage (stored, leveled, not used).
  \item \stat{INT} modifier to spell power (stored, leveled, not used).
  \item \stat{WIS} modifier to MP regen or spell accuracy.
  \item \stat{CHA} modifier to shop prices or NPC interactions.
  \item Per-class starting stat distributions (all classes share identical
        starting stats).
  \item Key items unlocking locked doors.
  \item Respawn mechanics (the \field{death} field is parsed but not acted on).
  \item Ring and amulet mechanical effects.
  \item Accuracy / miss chance (all attacks hit).
  \item Critical hits.
  \item Status conditions (\field{Conditions} field exists on
        \texttt{Character} and \texttt{EnemyInstance} but is never read).
\end{enumerate}

\end{document}
